---
title: "week4"
author: "Ilir_Sheraj"
date: "8/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The Big Short: Interest Rates Explained

```{r}
library(tidyverse)
```


```{r}
n <- 1000
loss_per_foreclosure <- -200000
p <- 0.02
defaults <- sample( c(0,1), n, prob=c(1-p, p), replace = TRUE)
sum(defaults * loss_per_foreclosure)
```

```{r}
B <- 10000
losses <- replicate(B, {
  defaults <- sample( c(0,1), n, prob=c(1-p, p), replace = TRUE) 
  sum(defaults * loss_per_foreclosure)
})
```

```{r}
data.frame(losses_in_millions = losses/10^6) %>%
    ggplot(aes(losses_in_millions)) +
    geom_histogram(binwidth = 0.6, col = "black")
```

```{r}
n*(p*loss_per_foreclosure + (1-p)*0)
```

```{r}
sqrt(n)*abs(loss_per_foreclosure)*sqrt(p*(1-p))
```

```{r}
x = - loss_per_foreclosure*p/(1-p)
x
x/180000
```

Abiut 2% interest rate to break even

Calculating interest rate for 1% probability of losing money

```{r}
l <- loss_per_foreclosure
z <- qnorm(0.01)
x <- -l*( n*p - z*sqrt(n*p*(1-p)))/ ( n*(1-p) + z*sqrt(n*p*(1-p)))
x/180000    # interest rate
loss_per_foreclosure*p + x*(1-p)    # expected value of the profit per loan
n*(loss_per_foreclosure*p + x*(1-p)) # expected value of the profit over n loans
```

```{r}
B <- 100000
profit <- replicate(B, {
    draws <- sample( c(x, loss_per_foreclosure), n, prob=c(1-p, p), replace = TRUE) 
    sum(draws)
})
mean(profit)    # expected value of the profit over n loans
mean(profit<0)    # probability of losing money
```

# The Big Short

Expected value with higher default rate and interest rate

```{r}
p <- .04
loss_per_foreclosure <- -200000
r <- 0.05
x <- r*180000
loss_per_foreclosure*p + x*(1-p)
```


Calculating number of loans for desired probability of losing money

```{r}
z <- qnorm(0.01)
l <- loss_per_foreclosure
n <- ceiling((z^2*(x-l)^2*p*(1-p))/(l*p + x*(1-p))^2)
n    # number of loans required
n*(loss_per_foreclosure*p + x * (1-p))    # expected profit over n loans
```

Monte Carlo simulation with known default probability

```{r}
B <- 10000
p <- 0.04
x <- 0.05 * 180000
profit <- replicate(B, {
    draws <- sample( c(x, loss_per_foreclosure), n, prob=c(1-p, p), replace = TRUE) 
    sum(draws)
})
mean(profit)
```

Monte Carlo simulation with unknown default probability

```{r}
p <- 0.04
x <- 0.05*180000
profit <- replicate(B, {
    new_p <- 0.04 + sample(seq(-0.01, 0.01, length = 100), 1)
    draws <- sample( c(x, loss_per_foreclosure), n, prob=c(1-new_p, new_p), replace = TRUE)
    sum(draws)
})
mean(profit)    # expected profit
mean(profit < 0)    # probability of losing money
mean(profit < -10000000)    # probability of losing over $10 million
```

```{r}
hist(profit)
```

# Assessment

Say you manage a bank that gives out 10,000 loans. The default rate is 0.03 and you lose $200,000 in each foreclosure. Create a random variable S that contains the earnings of your bank. Calculate the total amount of money lost in this scenario. Using the sample function, generate a vector called defaults that contains n samples from a vector of c(0,1), where 0 indicates a payment and 1 indicates a default. Multiply the total number of defaults by the loss per foreclosure.

```{r}
n <- 10000
loss_per_foreclosure <- -200000
p_default <- 0.03
set.seed(1)
defaults <- sample(c(0, 1), n, prob=c(1 - p_default, p_default), replace = TRUE)
S <- sum(defaults * loss_per_foreclosure)
S
```

Run a Monte Carlo simulation with 10,000 outcomes for S, the sum of losses over 10,000 loans. Make a histogram of the results.

```{r}
B <- 10000
n <- 10000
set.seed(1)
S <- replicate(B, {
  defaults <- sample( c(0,1), n, prob=c(1 - p_default, p_default), replace = TRUE) 
  sum(defaults * loss_per_foreclosure)
})
data.frame(losses = S) %>% ggplot(aes(losses)) +
  geom_histogram(binwidth = 1000000, color = "black")
```

What is the expected value of S, the sum of losses over 10,000 loans? For now, assume a bank makes no money if the loan is paid.

```{r}
n*(loss_per_foreclosure*p_default)
```

What is the standard error of S?

```{r}
sqrt(n)*abs(loss_per_foreclosure)*sqrt(p_default*(1-p_default))
```

So far, we've been assuming that we make no money when people pay their loans and we lose a lot of money when people default on their loans. Assume we give out loans for $180,000. How much money do we need to make when people pay their loans so that our net loss is $0? In other words, what interest rate do we need to charge in order to not lose money? If the amount of money lost or gained equals 0, the probability of default times the total loss per default equals the amount earned per probability of the loan being paid. Divide the total amount needed per loan by the loan amount to determine the interest rate.

```{r}
loss_per_foreclosure <- -200000
p_default <- 0.03
x <- -(loss_per_foreclosure*p_default) / (1 - p_default)
x/180000
```

With the interest rate calculated in the last example, we still lose money 50% of the time. What should the interest rate be so that the chance of losing money is 1 in 20? In math notation, what should the interest rate be so that Pr(S <0) = 0.05?

```{r}
n <- 10000
loss_per_foreclosure <- -200000
p_default <- 0.03

z <- qnorm(0.05)
x <- -loss_per_foreclosure*( n*p_default - z*sqrt(n*p_default*(1 - p_default)))/ ( n*(1 - p_default) + z*sqrt(n*p_default*(1 - p_default)))

x / 180000
```

# Week's Assignment

```{r}
library(tidyverse)
library(dslabs)
data(death_prob)
head(death_prob)
```

## Questions 1 and 2: Insurance rates, part 1

An insurance company offers a one-year term life insurance policy that pays 150,000 dollars in the event of death within one year. The premium (annual cost) for this policy for a 50 year old female is 1,150 dollars. Suppose that in the event of a claim, the company forfeits the premium and loses a total of 150,000, and if there is no claim the company gains the premium amount of $1,150. The company plans to sell 1,000 policies to this demographic.

The death_prob data frame from the dslabs package contains information about the estimated probability of death within 1 year (prob) for different ages and sexes. Use death_prob to determine the death probability of a 50 year old female, p.

```{r}
p <- death_prob %>% filter(age == 50 & sex == "Female") %>% select(prob) %>% unlist()
p
```

The loss in the event of the policy holder's death is -150,000 and the gain if the policy holder remains alive is the premium $1,150. What is the expected value of the company's net profit on one policy for a 50 year old female?

```{r}
1150*(1 - p) - 150000*p
```

Calculate the standard error of the profit on one policy for a 50 year old female.

```{r}
abs(1150 + 150000)*sqrt(p*(1-p))
```

What is the expected value of the company's profit over all 1,000 policies for 50 year old females?

```{r}
1000*(1150*(1 - p) - 150000*p)
```

What is the standard error of the sum of the expected value over all 1,000 policies for 50 year old females?

```{r}
abs(1150 + 150000)*sqrt(p*(1-p))*sqrt(1000)
```

Use the Central Limit Theorem to calculate the probability that the insurance company loses money on this set of 1,000 policies.

```{r}
pnorm(0, 667378.1, 269657.9)
```

50 year old males have a different probability of death than 50 year old females. We will calculate a profitable premium for 50 year old males in the following four-part question. 

Use death_prob to determine the probability of death within one year for a 50 year old male.

```{r}
male <- death_prob %>% filter(age == 50 & sex == "Male") %>% pull(prob)
male
```

Suppose the company wants its expected profits from 1,000 50 year old males with 150,000 life insurance policies to be $700,000. Use the formula for expected value of the sum of draws with the following values and solve for the premium. What premium should be charged?

```{r}
b <- ((700000/1000) - -150000*male)/(1-male)
b
```

Using the new 50 year old male premium rate, calculate the standard error of the sum of 1,000 premiums.

```{r}
err <- sqrt(1000) * abs(b - -150000) *sqrt(male*(1-male))
err
```

What is the probability of losing money on a series of 1,000 policies to 50 year old males?

```{r}
mu <- 1000*(b*(1 - male) - 150000*male)
pnorm(0, mu, err)
```

## Questions 3 and 4: insurance rates, part 2

Life insurance rates are calculated using mortality statistics from the recent past. They are priced such that companies are almost assured to profit as long as the probability of death remains similar. If an event occurs that changes the probability of death in a given age group, the company risks significant losses. In this 6-part question, we'll look at a scenario in which a lethal pandemic disease increases the probability of death within 1 year for a 50 year old to .015. Unable to predict the outbreak, the company has sold 1,000 $150,000 life insurance policies for 1,150.

What is the expected value of the company's profits over 1,000 policies?

```{r}
exp_val <- 1000*(1150*(1 - 0.015) - 150000*0.015)
exp_val
```

What is the standard error of the expected value of the company's profits over 1,000 policies?

```{r}
sd_err <- abs(1150 + 150000)*sqrt(0.015*(1-0.015))*sqrt(1000)
sd_err
```

What is the probability of the company losing money?

```{r}
pnorm(0, exp_val, sd_err)
```

Suppose the company can afford to sustain one-time losses of $1 million, but larger losses will force it to go out of business. What is the probability of losing more than 1 million?

```{r}
pnorm(-1000000, exp_val, sd_err)
```

Investigate death probabilities p <- seq(.01, .03, .001). What is the lowest death probability for which the chance of losing money exceeds 90%?

```{r}
p <- seq(.01, .03, .001)
money_loss <- sapply(p, function(x){
  mu <- 1000 * (-150000*x + 1150*(1-x))
  err <- sqrt(1000) * abs(-150000 - 1150) * sqrt(x*(1 - x))
  pnorm(0, mu, err)
})

min(p[which(money_loss > 0.9)])
```

Investigate death probabilities p <- seq(.01, .03, .0025). What is the lowest death probability for which the chance of losing over $1 million exceeds 90%?

```{r}
p <- seq(.01, .03, .0025)
money_loss <- sapply(p, function(x){
  mu <- 1000 * (-150000*x + 1150*(1-x))
  err <- sqrt(1000) * abs(-150000 - 1150) * sqrt(x*(1 - x))
  pnorm(-1000000, mu, err)
})

min(p[which(money_loss > 0.9)])
```

Question 4, which has two parts, continues the scenario from Question 3. Define a sampling model for simulating the total profit over 1,000 loans with probability of claim p_loss = .015, loss of -150,000 on a claim, and profit of $1,150 when there is no claim. Set the seed to 25, then run the model once. What is the reported profit (or loss) in millions (that is, divided by )?

```{r}
set.seed(25)
p <- .015
loss <- -150000
profit <- 1150
n <- 1000
outcomes <- sample(c(loss, profit), n, prob = c(p, 1-p), replace = TRUE)
sum(outcomes)/10^6
```

Set the seed to 27, then run a Monte Carlo simulation of your sampling model with 10,000 replicates to simulate the range of profits/losses over 1,000 loans.

```{r}
p <- 0.015
loss <- -150000
profit <- 1150
set.seed(27)
S <- replicate(10000, {
  outcomes <- sample(c(loss, profit), 1000, prob = c(p, 1-p), replace = TRUE)
  sum(outcomes)/10^6
})
mean(S< -1)
```

## Questions 5 and 6: Insurance rates, part 3

Question 5, which has 4 parts, continues the pandemic scenario from Questions 3 and 4. Suppose that there is a massive demand for life insurance due to the pandemic, and the company wants to find a premium cost for which the probability of losing money is under 5%, assuming the death rate stays stable at p = 0.015. 
Calculate the premium required for a 5% chance of losing money given n = 1000 loans probability of death p = 0.015, and loss per claim l=-150000. Save this premium as x for use in further questions.

```{r}
p <- 0.015
n <- 1000
l <- -150000
z <- qnorm(0.05)
x <- -l*( n*p - z*sqrt(n*p*(1-p)))/ ( n*(1-p) + z*sqrt(n*p*(1-p)))
x
```

What is the expected profit per policy at this rate?

```{r}
l*p + x*(1-p)
```

What is the expected profit over 1,000 policies?

```{r}
1000*(l*p + x*(1-p))
```

Run a Monte Carlo simulation with B=10000 to determine the probability of losing money on 1,000 policies given the new premium x, loss on a claim of $150,000, and probability of claim p = 0.015. Set the seed to 28 before running your simulation. What is the probability of losing money here?

```{r}
set.seed(28)
B <- 10000
profit <- replicate(B, {
  draws <- sample(c(x, l), n, prob=c(1-p, p), replace = TRUE)
  sum(draws)
})
mean(profit < 0)
```

The company cannot predict whether the pandemic death rate will stay stable. Set the seed to 29, then write a Monte Carlo simulation that for each of B=10000 iterations. randomly changes p by adding a value between -0.01 and 0.01 with sample(seq(-0.01, 0.01, length = 100), 1). use the new random p to generate a sample of n=1000 policies with premium x and loss per claim l=-150000. returns the profit over n policies (sum of random variable) The outcome should be a vector of B total profits.

```{r}
set.seed(29)
n <- 1000
B <- 10000
l <- -150000
p <- 0.015
X <- replicate(B, {
  new_p <- p + sample(seq(-0.01, 0.01, length=100), 1)
  Y <- sample(c(x, l), n, replace=TRUE, prob=c(1-new_p, new_p))
  sum(Y)
})
```

What is the expected value over 1,000 policies?

```{r}
set.seed(29, sample.kind="Rounding")
n <- 1000
B <- 10000
l <- -150000
p <- 0.015
profit <- replicate(B, {
    new_p <- p + sample(seq(-0.01, 0.01, length = 100), 1)
    draws <- sample(c(x, l), n, 
                    prob=c(1-new_p, new_p), replace = TRUE) 
    sum(draws)
})
mean(profit)
```

What is the probability of losing money?

```{r}
mean(profit < 0)
```

What is the probability of losing more than $1 million?

```{r}
mean(profit < -1000000)
```

